---
title: "CBASS"
author: "Jill Ashey"
date: "2024-12-13"
output: html_document
---

This script uses the [CBASSED50 R package](https://github.com/reefgenomics/CBASSED50/tree/main) to model ED50 values from CBASS experiments. The following CBASS experiments were done at the CBASS workshop at the Maldives in December 2024. I am using the CBASS [demo script](https://github.com/reefgenomics/CBASSED50/blob/main/CBASSED50_demo.qmd) as a model for this script (I just wanted to write it myself so it would make more sense to me). 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(CBASSED50)
library(dplyr)
library(tidyr)
library(ggplot2)
library(readxl)
library(rstudioapi)
library(RColorBrewer)
```

Read in data 
```{r}
cbass_data <- read.csv("../data/cbass/RAW_20241212_MV_MarHE_CBASSED50.csv")
```

Make data tidy and validate. 
```{r process-and-validate-cbass-dataset}
#cbass_data <- preprocess_dataset(cbass_data)
validate_cbass_dataset(cbass_data)
```
Need to look into what `preprocess_dataset` is doing...

For the ED50 model, a grouping property needs to be chosen; in our case, it is Species (Poc, Por), and Condition (shallow or deep). Genotype should NEVER be used as a grouping property when using the CBASS R package, as this column is used as a `curveid` argument for `drm` modeling. 

Fit dose response models 
```{r}
grouping_properties <- c("Site", "Condition", "Species", "Timepoint")
drm_formula <- "PAM ~ Temperature"
models <- fit_drms(cbass_data, grouping_properties, drm_formula, is_curveid = TRUE)
```

Get ED5s, ED50s and ED95s from models 
```{r}
eds <- get_all_ed_by_grouping_property(models)
cbass_data <- define_grouping_property(cbass_data, grouping_properties) %>%
  mutate(GroupingProperty = paste(GroupingProperty, Genotype, sep = "_"))

eds_df <- 
  left_join(eds, cbass_data, by = "GroupingProperty") %>%
  select(names(eds), all_of(grouping_properties)) %>%
  distinct()

# Save as csv 
write.csv(eds_df, "../output/cbass/EDs.csv", row.names = F)
```

Make ED boxplots 
```{r}
eds_boxplot <- eds_df %>% ggplot(
  aes(x = Species, y = ED50, color = Condition)) +
  geom_boxplot() + 
  stat_summary(
    fun = mean, 
    geom = "text", 
    aes(label = round(after_stat(y), 2)), show.legend = F,
    position = position_dodge(width = 0.75),
    vjust = -1
  ) +
  facet_grid(~ Site) +
  ylab("ED50s - Temperatures [C°]")+
  scale_color_brewer(palette = "Set2"); eds_boxplot

ggsave("../output/cbass/ED50_boxplot.pdf", plot = eds_boxplot)
ggsave("../output/cbass/ED50_boxplot.png", plot = eds_boxplot)
```

Check out curves by genotype 
```{r}
exploratory_curve <-
  ggplot(data = cbass_data,
       aes(
         x = Temperature,
         y = PAM,
         # You can play around with the group value (e.g., Species, Site, Condition)
         group = GroupingProperty,
         color = Genotype)) +
  geom_smooth(
    method = drc::drm,
    method.args = list(
      fct = drc::LL.3()),
    se = FALSE,
    size = 0.7
  ) +
  geom_point(size = 1.5) +
  facet_grid(Species ~ Site ~ Condition); exploratory_curve
```

Predict PAM values for temperature range in CBASS experiment 
```{r}
# First fit models without curveid
models <- fit_drms(cbass_data, grouping_properties, drm_formula, is_curveid = FALSE)
# The default number of values for range of temperatures is 100
temp_ranges <- define_temperature_ranges(cbass_data$Temperature, n=100)
predictions <- get_predicted_pam_values(models, temp_ranges)
```

NAs may be produced, which happens when PAM at a higher temperature exceed PAM values at a lower temperature (the model assumes decreasing PAM values with increasing temperatures).

Tidy data for visualization
```{r}
predictions_df <- 
  left_join(predictions,
            define_grouping_property(cbass_data, grouping_properties) %>% 
              select(c(all_of(grouping_properties), GroupingProperty)),
            by = "GroupingProperty",
            relationship = "many-to-many") %>%
  distinct()
```

Get ED summary stats for groupings 
```{r}
summary_eds_df <- eds_df %>%
  group_by(Site, Condition, Species, Timepoint) %>%
    summarise(Mean_ED5 = mean(ED5),
            SD_ED5 = sd(ED5),
            SE_ED5 = sd(ED5) / sqrt(n()),
            Conf_Int_5 = qt(0.975, df = n() - 1) * SE_ED5,
            Mean_ED50 = mean(ED50),
            SD_ED50 = sd(ED50),
            SE_ED50 = sd(ED50) / sqrt(n()),
            Conf_Int_50 = qt(0.975, df = n() - 1) * SE_ED50,
            Mean_ED95 = mean(ED95),
            SD_ED95 = sd(ED95),
            SE_ED95 = sd(ED95) / sqrt(n()),
            # The value 0.975 corresponds to the upper tail probability
            # for a two-tailed t-distribution with a 95% 
            Conf_Int_95 = qt(0.975, df = n() - 1) * SE_ED95) %>%
  mutate(across(c(Mean_ED50, SD_ED50, SE_ED50,
                  Mean_ED5, SD_ED5, SE_ED5,
                  Mean_ED95, SD_ED95, SE_ED95,
                  Conf_Int_5,Conf_Int_50,Conf_Int_95), ~round(., 2)))

# Save as csv 
write.csv(summary_eds_df, "../output/cbass/summaryEDs.csv", row.names = F)
```

Join all data 
```{r}
result_df <- predictions_df %>%
  left_join(summary_eds_df, by = c("Site", "Condition", "Species", "Timepoint"))
```

Plot EDs for groups 
```{r}
tempresp_curve <- ggplot(result_df,
       aes(x = Temperature,
           y = PredictedPAM,
           #group = GroupingProperty,
           # You can customize the group here
           color = Condition)) +
  geom_line() +
  geom_ribbon(aes(ymin = Upper,
                  ymax = Lower,
                  fill = Condition),
              alpha = 0.2,
              linetype = "dashed") +
  geom_segment(aes(x = Mean_ED5,
                   y = 0,
                   xend = Mean_ED5,
                   yend = max(Upper)),
               linetype = 3) +
  geom_text(mapping=aes(x = Mean_ED5,
                        y = max(Upper) + 0.12,
                        label = round(Mean_ED5, 2)),
            size = 3, angle = 90, check_overlap = T) +
  geom_segment(aes(x = Mean_ED50,
                   y = 0,
                   xend = Mean_ED50,
                   yend = max(Upper)),
               linetype = 3) +
  geom_text(mapping=aes(x = Mean_ED50,
                        y = max(Upper) + 0.12,
                        label = round(Mean_ED50, 2)),
            size = 3, angle = 90, check_overlap = T) +
  geom_segment(aes(x = Mean_ED95,
                   y = 0,
                   xend = Mean_ED95,
                   yend = max(Upper)),
               linetype = 3) +
  geom_text(mapping=aes(x = Mean_ED95,
                        y = max(Upper) + 0.12,
                        label = round(Mean_ED95, 2)),
            size = 3, angle = 90, check_overlap = T) +
  facet_grid(Species ~ Site ~ Condition) +
  # To add the real PAM and compare with predicted values
  geom_point(data = cbass_data,
             aes(x = Temperature,
                 y = PAM)) +
                 xlab("Temperature [C°]")+
  scale_y_continuous(expand = c(0, 0.2)); tempresp_curve

ggsave("../output/cbass/tempresp_curve.pdf", plot = tempresp_curve)
ggsave("../output/cbass/tempresp_curve.png", plot = tempresp_curve)
```





