---
title: "cbass statistics"
author: "Jill Ashey"
date: "2024-12-15"
output: html_document
---

The following script assess statistical significance between the ED50 values of four coral species in the Maldives

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(broom)
library(ggstatsplot)
```

Read in data 
```{r}
cbass <- read.csv("../output/cbass/EDs_ALL.TAXA.csv")
```

Calculate the decline width (DW), which is: ED95 - ED5, along with relative thermal tolerance (RTT), which is: ED50 - MMM (in our case, MMM is 29.5).
```{r}
cbass <- cbass %>%
  mutate(DW = ED95 - ED5) %>%
  mutate(RTT = ED50 - 29.5)
```

Compute summary statistics per species 
```{r}
summary_stats <- cbass %>%
  group_by(Species) %>%
  summarise(
    mean_ED5 = mean(ED5),
    sd_ED5 = sd(ED5),
    ci_lower_ED5 = mean_ED5 - qt(0.975, df=n()-1) * (sd_ED5 / sqrt(n())),
    ci_upper_ED5 = mean_ED5 + qt(0.975, df=n()-1) * (sd_ED5 / sqrt(n())),
    
    mean_ED50 = mean(ED50),
    sd_ED50 = sd(ED50),
    ci_lower_ED50 = mean_ED50 - qt(0.975, df=n()-1) * (sd_ED50 / sqrt(n())),
    ci_upper_ED50 = mean_ED50 + qt(0.975, df=n()-1) * (sd_ED50 / sqrt(n())),
    
    mean_ED95 = mean(ED95),
    sd_ED95 = sd(ED95),
    ci_lower_ED95 = mean_ED95 - qt(0.975, df=n()-1) * (sd_ED95 / sqrt(n())),
    ci_upper_ED95 = mean_ED95 + qt(0.975, df=n()-1) * (sd_ED95 / sqrt(n())),
    
    mean_DW = mean(DW),
    sd_DW = sd(DW),
    ci_lower_DW = mean_DW - qt(0.975, df=n()-1) * (sd_DW / sqrt(n())),
    ci_upper_DW = mean_DW + qt(0.975, df=n()-1) * (sd_DW / sqrt(n())),
    
    mean_RTT = mean(RTT),
    sd_RTT = sd(RTT),
    ci_lower_RTT = mean_RTT - qt(0.975, df=n()-1) * (sd_RTT / sqrt(n())),
    ci_upper_RTT = mean_RTT + qt(0.975, df=n()-1) * (sd_RTT / sqrt(n()))
  )
```

Run ANOVA to compare species ED50s
```{r}
# Run ANOVA
anova_result <- aov(ED50 ~ Species, data = cbass)

# Print summary of the ANOVA
summary(anova_result)

# Check normality
shapiro.test(residuals(anova_result))

# Check homogeneity of variances (install 'car' package if not installed)
leveneTest(ED50 ~ Species, data = cbass)
```
Data is normal and ED50 is significant by species. 

Run post-hoc tests
```{r}
# Run Tukey's HSD post-hoc test
tukey_result <- TukeyHSD(anova_result)

# Print the results of Tukey's HSD
print(tukey_result)
```

Significant differences between POC annd POR (still need to add acropora). 

Plot boxplot of ED50
```{r}
# Boxplot of ED50 by Species
ggplot(cbass, aes(x = Species, y = ED50)) +
    geom_boxplot(fill = "lightblue", color = "darkblue") +
    theme_minimal() +
    labs(title = "Boxplot of ED50 by Species", x = "Species", y = "ED50")

```

Run ANOVA to compare species ED5s
```{r}
# Run ANOVA
anova_result <- aov(ED5 ~ Species, data = cbass)

# Print summary of the ANOVA
summary(anova_result)

# Check normality
shapiro.test(residuals(anova_result))

# Check homogeneity of variances (install 'car' package if not installed)
leveneTest(ED5 ~ Species, data = cbass)
```

Data is normal and ED5 is significant by species. 

Run post-hoc tests
```{r}
# Run Tukey's HSD post-hoc test
tukey_result <- TukeyHSD(anova_result)

# Print the results of Tukey's HSD
print(tukey_result)
```

Plot boxplot of ED5
```{r}
# Boxplot of ED50 by Species
ggplot(cbass, aes(x = Species, y = ED5)) +
    geom_boxplot(fill = "lightblue", color = "darkblue") +
    theme_minimal() +
    labs(title = "Boxplot of ED50 by Species", x = "Species", y = "ED5")
```

Run ANOVA to compare species ED95s
```{r}
# Run ANOVA
anova_result <- aov(ED95 ~ Species, data = cbass)

# Print summary of the ANOVA
summary(anova_result)

# Check normality
shapiro.test(residuals(anova_result))

# Check homogeneity of variances (install 'car' package if not installed)
leveneTest(ED95 ~ Species, data = cbass)
```

Data is normal and ED95 is significant by species. 

Run post-hoc tests
```{r}
# Run Tukey's HSD post-hoc test
tukey_result <- TukeyHSD(anova_result)

# Print the results of Tukey's HSD
print(tukey_result)
```

Plot boxplot of ED95
```{r}
# Boxplot of ED95 by Species
ggplot(cbass, aes(x = Species, y = ED95)) +
    geom_boxplot(fill = "lightblue", color = "darkblue") +
    theme_minimal() +
    labs(title = "Boxplot of ED95 by Species", x = "Species", y = "ED95")
```

Run ANOVA to compare species DWs
```{r}
# Run ANOVA
anova_result <- aov(DW ~ Species, data = cbass)

# Print summary of the ANOVA
summary(anova_result)

# Check normality
shapiro.test(residuals(anova_result))

# Check homogeneity of variances (install 'car' package if not installed)
leveneTest(DW ~ Species, data = cbass)
```

Run post-hoc tests
```{r}
# Run Tukey's HSD post-hoc test
tukey_result <- TukeyHSD(anova_result)

# Print the results of Tukey's HSD
print(tukey_result)
```

Plot boxplot of DW
```{r}
# Boxplot of ED95 by Species
ggplot(cbass, aes(x = Species, y = DW)) +
    geom_boxplot(fill = "lightblue", color = "darkblue") +
    theme_minimal() +
    labs(title = "Boxplot of DW by Species", x = "Species", y = "DW")
```

Run ANOVA to compare species RTTs
```{r}
# Run ANOVA
anova_result <- aov(RTT ~ Species, data = cbass)

# Print summary of the ANOVA
summary(anova_result)

# Check normality
shapiro.test(residuals(anova_result))

# Check homogeneity of variances (install 'car' package if not installed)
leveneTest(RTT ~ Species, data = cbass)
```

Run post-hoc tests
```{r}
# Run Tukey's HSD post-hoc test
tukey_result <- TukeyHSD(anova_result)

# Print the results of Tukey's HSD
print(tukey_result)
```

Plot boxplot of RTT
```{r}
# Boxplot of ED95 by Species
ggplot(cbass, aes(x = Species, y = RTT)) +
    geom_boxplot(fill = "lightblue", color = "darkblue") +
    theme_minimal() +
    labs(title = "Boxplot of RTT by Species", x = "Species", y = "RTT")
```





